# Data acquisition and extraction

## Access protocols and permissions

* Reproducible extraction of data from source location: may be complicated by access protocols.

  - access tokens; APIs
  - raw data from github for private repos
  - databases
  - package `httr` to access data from websites


```
---- Forwarded Message -----
From: GitHub <noreply@github.com>
To: Arthur Small <asmall@virginia.edu>
Sent: Sunday, February 21, 2021, 6:20:58 AM EST
Subject: [GitHub] Deprecation Notice

Hi @arthursmalliii,

You recently used a password to access the repository at uva-eng-time-series-sp21/coronato-nicholas with git using git/2.30.0.

Basic authentication using a password to Git is deprecated and will soon no longer work. Visit https://github.blog/2020-12-15-token-authentication-requirements-for-git-operations/ for more information around suggested workarounds and removal dates.

Thanks,
The GitHub Team
```



## Accessing databases

```{r open connection to database 2, eval=FALSE, include=FALSE}
# Open connection to a remote database
# Make sure your VPN network connection is active if needed!

# if(!('RPostgreSQL' %in% installed.packages())) install.packages('RPostgreSQL')
library(RPostgreSQL)

# file "my_postgres_credentials.R" contains the log-in information
# DON'T EVER HARD CODE ACCESS CREDENTIALS INTO YOUR SOURCE CODE!!!
source("my_postgres_db_credentials.R")


# Open connection
db_driver <- dbDriver("PostgreSQL")
db <- dbConnect(db_driver,user=user, password=password,dbname="postgres", host=host)
rm(password) 

# check the connection: If function returns value TRUE, the connection is working
dbExistsTable(db, "metadata")
```

```{r retrieve data from db 2, eval=FALSE, message=FALSE}
esales <- dbGetQuery(db,'SELECT * from eia_elec_sales_va_all_m') # SQL code to retrieve data from a table in the remote database
# str(esales)
esales <- as_tibble(esales) # Convert dataframe to a 'tibble' for tidyverse work
# str(esales)
```

```{r save data in Apache Arrow format 2, eval=FALSE}
# Reference: https://arrow.apache.org/docs/r/
# if(!('arrow' %in% installed.packages())) install.packages('arrow')
library(arrow)
write_feather(esales, "esales.feather")
```

```{r close db connection 2, eval=FALSE}
# Close connection -- this is good practice
dbDisconnect(db)
dbUnloadDriver(db_driver)
```


## Other comments

Make your extraction code "as reproducible as possible", subject to these access constraints. At minimum, document clearly how you obtained the data, so others could follow your path, even if not via pure code.
 
Keep your raw data in read-only mode. Don't edit these files. 

Write code to transform the raw data into form you will use for analysis. Don't do it manually.


