# Transform a data frame to `tsibble` object

Readings: FPP3, Section 2.1

Convert the data frame into a time series `tsibble` object.

```{r}
# install.packages("tsibble")
library(tsibble) # Reference: https://tsibble.tidyverts.org/articles/intro-tsibble.html

esales <- arrow::read_feather("data/esales.feather")
esales %>%
  dplyr::select(date, sales_GWh = value) -> esales_tbl
esales_tbl %>% as_tsibble(index = date) -> elsales_tbl_ts

print(elsales_tbl_ts)
```


## Time indexing

See R4DS ch. 16.

Depending on how dates and times are recorded in your raw data, you may face more or less work to organize them into form(s) suitable as `tsibble` index variable.


The `lubridate` and `hms` packages may be valuable.

```{r Change the tsibble to index monthly}
# install.packages("feasts"), Reference: https://feasts.tidyverts.org/
library(feasts)

elsales_tbl_ts %>%
  mutate(Month = tsibble::yearmonth(date)) %>%
  as_tsibble(index = Month) %>%
  dplyr::select(Month,sales_GWh) -> vaelsales_tbl_ts

print(vaelsales_tbl_ts)
```


## Running diagnostics on your tsibble

** Ideally, should have exactly one row (i.e., one vector of measured values) for each time interval (`index`) and each value of the `key` variables. 
 -- May not have any duplicates.
 -- May have missing values
 
### Duplicate values 


### Missing values


### Irregular time intervals
