# (PART) Exploratory Data Analysis {-} 

# Exploratory analysis of time series data

```{r set up coding environment, include=FALSE, message=FALSE}
# library(dplyr) -- don't need this if you are loading the entire 'tidyverse' suite
library(tidyverse)
library(lubridate) # For easy handling of time-indexed objects

# if(!('fpp3' %in% installed.packages())) install.packages('fpp3')
library(fpp3)
```

## Overview

Readings: 

* FPP3, Sections 2.2--2.6. 
* [TSDS, Section 3.10](https://jhudatascience.org/tidyversecourse/wrangle-data.html#exploratory-data-analysis), [Chapter 4](https://jhudatascience.org/tidyversecourse/dataviz.html) and [Section 5.5](https://jhudatascience.org/tidyversecourse/model.html#descriptive-and-exploratory-analysis)

Topics:

* Time series plots.
* Trends. Seasonal (periodic) patterns. Cycles.
* Seasonal plots. Seasonal sub-series.
* Investigating relationships between two variables. Scatterplots. Correlation. Scatterplot matrices.

Assignment: Explore your data.


## Briefly characterize the dataset

Provide a brief example of the data, showing how they are structured.

Example: Monthly electricity sales for Virginia

Previously we extracted monthly electricity sales data for Virginia from a remote database, converted the data frame into a `tibble` object, and saved the result to a file in feather format.


```{r read in data}
library(arrow)
esales <- read_feather("data/esales.feather")
```

```{r make basic data summaries}
print(esales)    # print the data as a table
summary(esales)  # compute basic summary statistics about the data
boxplot(esales)
hist(elsales_tbl_ts$sales_GWh, breaks=40) #  Make a histogram of monthly sales
```


### Examine subsets of the data

```{r use tidyverse syntax to perform some simple data manipulations, eval=FALSE}
# References: https://www.tidyverse.org/, https://dplyr.tidyverse.org/
# filter(data object, condition) : syntax for filter() command

esales %>%
  filter(year == 2019) %>%
  filter(value > 9000) %>%
  print()

(esales %>%
  group_by(year) %>%
  summarise(Total = sum(value)) -> total_esales_by_year)

esales %>%
  mutate(sales_TWh = value/1000) %>%
  dplyr::select(-value)
```


```{r}
# library(lubridate) # Make it easy to deal with dates

esales %>% filter(month==3)                   # These three lines of code
esales %>% filter(month(date)==3)             #   all do
esales %>% filter(lubridate::month(date)==3)  #   the same thing.

# We don't have to keep the 'year' and 'month' column: can recover them if needed

esales %>%
  dplyr::select(date, sales_GWh = value) -> esales_tbl

print(esales_tbl)
```


## Plot the time series

Ref: FPP3, Section 2.2


```{r use ggplot2 to generate a plot}
#Reference: https://ggplot2.tidyverse.org/

ggplot(data=esales, aes(x=date,y=value)) +
  geom_line() + xlab("Year") + ylab("Virginia monthly total electricity sales (GWh)")
```

```{r use feasts autoplot}
# feasts::autoplot() is handy for quickly generating time series plots

autoplot(vaelsales_tbl_ts, sales_GWh) +
  ylab("Virginia monthly total electricity sales (GWh)") +
  xlab("")  # Leave horiz. axis label blank
```


## Sesaonal plots

Ref: FPP3, Sections 2.3, 2.4

### Example: Virginia monthly electricity

Recall how we readied these data:

```{r}
esales <- arrow::read_feather("data/esales.feather")
esales %>%
  dplyr::select(date, sales_GWh = value) -> esales_tbl

esales_tbl %>% as_tsibble(index = date) -> elsales_tbl_ts

print(elsales_tbl_ts)
```


```{r, eval=FALSE}
# This plot won't work. Why not?
elsales_tbl_ts %>%
  feasts::gg_season(sales_GWh, labels = "both") + ylab("Virginia electricity sales (GWh)")
```

```{r Change the tsibble so index is monthly}
# install.packages("feasts"), Reference: https://feasts.tidyverts.org/
library(feasts)

elsales_tbl_ts %>%
  mutate(Month = tsibble::yearmonth(date)) %>%
  as_tsibble(index = Month) %>%
  dplyr::select(Month,sales_GWh) -> vaelsales_tbl_ts

print(vaelsales_tbl_ts)
```


```{r , warning=FALSE}
vaelsales_tbl_ts %>% gg_season(sales_GWh, labels = "both") + ylab("Virginia electricity sales (GWh)")
```

### Example: Australian production

```{r, warning=FALSE}
# install.packages('tsibbledata')
library(tsibbledata)

aus_production

aus_production %>% gg_season(Electricity)

aus_production %>% gg_season(Beer)
```


### Seasonal subseries plots

Ref: FPP3, Section 2.5

```{r}
vaelsales_tbl_ts %>%
  gg_subseries(sales_GWh) +
  labs(
    y = "Sales (GWh)",
    title = "Seasonal subseries plot: Virginia electricity sales"
  )
```


## Scatterplots

Readings: FPP Sect. 2.6

Investigating relationships between two variables. Scatterplots. Correlation. Scatterplot matrices.

```{r}
vic_elec

summary(vic_elec)

vic_elec %>%
  filter(year(Time) == 2013) %>%
  autoplot(Demand) +
  labs(
    y = "Demand (GW)",
    title = "Half-hourly electricity demand: Victoria"
  )
```

```{r}
vic_elec %>%
  filter(year(Time) == 2013) %>%
  autoplot(Temperature) +
  labs(
    y = "Temperature (degrees Celsius)",
    title = "Half-hourly temperatures: Melbourne, Australia"
  )
```

```{r}
vic_elec %>%
  filter(year(Time) == 2013) %>%
  ggplot(aes(x = Temperature, y = Demand)) +
#  geom_density2d() +
  geom_point(size=0.1, aes(colour=Holiday), alpha = 0.4) +
  labs(y = "Demand (GW)", x = "Temperature (degrees Celsius)")
```

### A Scatterplot matrix

```{r, warning=FALSE}
vic_elec

boxplot(vic_elec$Temperature)

# install.packages("GGally")
vic_elec %>%
  # mutate(Temperature = round(Temperature)) %>%
  # pivot_wider(values_from=c(Demand,Temperature), names_from=Holiday) %>%
  GGally::ggpairs(columns = 3:2)

vic_elec %>%
  mutate(Year = factor(year(Date))) %>%
  dplyr::select(-c(Date, Holiday)) %>%
  GGally::ggpairs(columns = 4:2)
```


