# Time series decomposition

Readings: FPP3, Sections 3.1, 3.2


### Example: Gross Domestic Product data


```{r}
library(tsibbledata) # Data sets package

print(global_economy)
```

```{r}
global_economy %>% filter(Country=="Sweden") %>% print()
```

```{r}
global_economy %>%
  filter(Country=="Sweden") %>%
  autoplot(GDP) +
  ggtitle("GDP for Sweden") + ylab("$US billions")
```



### Fitting data to simple models

```{r}
global_economy %>% model(trend_model = TSLM(GDP ~ trend())) -> fit

fit
```

```{r}
fit %>% filter(Country == "Sweden") %>% residuals()
```

```{r}

fit %>% filter(Country == "Sweden") %>% residuals() %>% autoplot(.resid)
```

### Work with ln(GDP)

```{r}
global_economy %>%
  filter(Country=="Sweden") %>%
  autoplot(log(GDP)) +
  ggtitle("ln(GDP) for Sweden") + ylab("$US billions")
```

```{r}
global_economy %>%
  model(trend_model = TSLM(log(GDP) ~ trend())) -> logfit
```

```{r}
logfit %>% filter(Country == "Sweden") %>% residuals() %>% autoplot()
```

```{r}
global_economy %>% model(trend_model = TSLM(log(GDP) ~ log(Population))) -> fit3

fit3 %>% filter(Country == "Sweden") %>% residuals() %>% autoplot()

```

## Producing forecasts

```{r}
fit %>% forecast(h = "3 years") -> fcast3yrs

fcast3yrs

```

```{r}

fcast3yrs %>% filter(Country == "Sweden", Year == 2020) %>% str()
```

```{r visualize forecasts}
fcast3yrs %>%
  filter(Country=="Sweden") %>%
  autoplot(global_economy) +
  ggtitle("GDP for Sweden") + ylab("$US billions")
```

### Model residuals vs. forecast errors

Model residuals:

Your data: $y_1, y_2, \ldots, y_T$

Fitted values: $\hat{y}_1, \hat{y}_2, \ldots, \hat{y}_T$

Model residuals: $e_t = y_t - \hat{y}_t$

Forecast errors:

```{r}
augment(fit)
```

```{r}
augment(fit) %>% filter(Country == "Sweden") %>%
  ggplot(aes(x = .resid)) +
  geom_histogram(bins = 20) +
  ggtitle("Histogram of residuals")
```

### Are the model residuals auto-correlated?

```{r}
augment(fit) %>% filter(Country == "Sweden") -> augSweden

augSweden %>%
  ACF(.resid) %>%
  autoplot() + ggtitle("ACF of residuals")
```

```{r}
augment(fit3) %>% filter(Country == "Sweden") -> augSweden3

augSweden3 %>%
  ACF(.resid) %>%
  autoplot() + ggtitle("ACF of residuals")
```


## Example: GDP, several countries


```{r}
library(tsibbledata) # Data sets package

nordic <- c("Sweden", "Denmark", "Norway", "Finland")

(global_economy %>% filter(Country %in% nordic) -> nordic_economy)

```

```{r}
nordic_economy %>% autoplot(GDP)
```

```{r}
fitnord <- nordic_economy %>%
  model(
    trend_model = TSLM(GDP ~ trend()),
    trend_model_ln = TSLM(log(GDP) ~ trend()),
    ets = ETS(GDP ~ trend("A")),
    arima = ARIMA(GDP)
  )

fitnord
```

```{r}
fitnord %>%
  dplyr::select(arima) %>%
  coef()
```


Denmark: ARMA(1,1)

Finland: MA(2)

Norway: MA(1)

Sweden: MA(2)

```{r}
nordic_economy %>%
  model(arima_constrained = ARIMA(GDP ~ pdq(1,0,2))) %>% dplyr::select(arima_constrained) %>% coef()
```

```{r}
fitnord %>% coef()
```

```{r}
fitnord %>%  glance()  
```

```{r}
fitnord %>% filter(Country == "Denmark") %>% dplyr::select(arima) %>% report()
```

```{r}
fitnord %>%
  accuracy() %>%
  arrange(Country, MPE)
```



```{r, eval=FALSE}
# ETS forecasts
USAccDeaths %>%
  ets() %>%
  forecast() %>%
  autoplot()
```

```{r, eval=FALSE}
str(taylor)
plot(taylor)
```

### Plot lagged values

```{r plot lagged values}
vaelsales_tbl_ts  %>% filter(month(Month) %in% c(3,6,9,12)) %>% gg_lag(sales_GWh, lags = 1:2)

vaelsales_tbl_ts  %>% filter(month(Month) == 1) %>% gg_lag(sales_GWh, lags = 1:2)
```

```{r}
vaelsales_tbl_ts %>% ACF(sales_GWh) %>% autoplot()
```

```{r perform automated time series decomposition}


# decompose(vaelsales_tbl_ts)
```

```{r perform additive STL decomposition of the VA electricity sales time series}
vaelsales_tbl_ts %>%
  model(STL(sales_GWh ~ trend(window=21) + season(window='periodic'), robust = TRUE)) %>%
  components() %>%
  autoplot()
```

```{r perform multiplicative STL decomposition of the VA electricity sales time series}
vaelsales_tbl_ts %>%
  mutate(ln_sales_GWh = log(sales_GWh)) %>%
  model(STL(ln_sales_GWh ~ trend(window=21) + season(window='periodic'),
    robust = TRUE)) %>%
  components() %>%
  autoplot()
```

```{r}
vaelsales_tbl_ts %>%
  features(sales_GWh, feat_stl)
```

```{r}
vaelsales_tbl_ts %>%
  features(sales_GWh, feature_set(pkgs="feasts"))
```
